---
title: "Bulk RNA-seq Analysis Report"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    format-links: false
    embed-resources: true
    code-fold: true
    theme: zephyr
    css: assets/style.css
    include-after-body: assets/report-ui.html
    fig-align: center
    fig-responsive: true
  pdf:
    toc: true
    toc-depth: 3
    number-sections: false
    fig-align: center
    pdf-engine: xelatex
    geometry:
      - margin=0.65in
    colorlinks: true
    documentclass: article
    papersize: letter
knitr:
  opts_chunk:
    dev: "png"
    dpi: 110
    fig.retina: 2
    out.width: "100%"
params:
  config_path: "config/config.yaml"
  project_root: "."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

# Allow reused chunk labels inside per-contrast knit_child loops;
# knitr auto-suffixes so figure files become <label>-1.png, <label>-2.png, ...
options(knitr.duplicate.label = "allow")

suppressPackageStartupMessages({
  library(yaml)
  library(readr)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(DT)
  library(plotly)
  library(jsonlite)
})

`%||%` <- function(a, b) if (is.null(a)) b else a

project_root <- normalizePath(params$project_root, mustWork = TRUE)

resolve_project_path <- function(path) {
  if (is.null(path) || is.na(path) || !nzchar(path)) {
    stop("Required path parameter is empty.")
  }
  if (!grepl("^(/|~|[A-Za-z]:[/\\\\])", path) && !file.exists(path)) {
    path <- file.path(project_root, path)
  }
  normalizePath(path, winslash = "/", mustWork = TRUE)
}

config_path <- resolve_project_path(params$config_path)

# All downstream paths are resolved relative to the project root.
rp <- function(...) file.path(project_root, ...)

cfg      <- yaml::read_yaml(config_path)
project  <- cfg$project       %||% list()
exp_meta <- project$experiment %||% list()

samples_tsv   <- rp(cfg$input$samples_tsv)
contrasts_tsv <- rp(cfg$input$contrasts_tsv)

samples   <- readr::read_tsv(samples_tsv,   show_col_types = FALSE)
contrasts <- readr::read_tsv(contrasts_tsv, show_col_types = FALSE)

# Helper to look up contrast metadata by id.
contrast_row <- function(cid) {
  out <- contrasts[contrasts$contrast_id == cid, , drop = FALSE]
  if (nrow(out) == 0) return(NULL)
  as.list(out[1, ])
}

# Shared helper: emit "no results available" card.
no_results <- function(msg = "No results available for this section.") {
  htmltools::div(
    class  = "callout callout-note",
    style  = "padding:0.75em 1em;border-left:4px solid #8aa;background:#f5f7f9;margin:0.5em 0;",
    htmltools::tags$em(msg)
  )
}

# Safe CSV/TSV loader: returns NULL on missing/unreadable/empty.
safe_read <- function(path, sep = ",") {
  if (is.null(path) || !file.exists(path)) return(NULL)
  df <- tryCatch(
    if (sep == "\t") {
      readr::read_tsv(path, show_col_types = FALSE)
    } else {
      readr::read_csv(path, show_col_types = FALSE)
    },
    error = function(e) NULL
  )
  if (is.null(df) || nrow(df) == 0) return(df)  # keep empty DF for header-only files
  df
}

fmt_num <- function(x, digits = 3) {
  if (is.null(x) || length(x) == 0) return(NA_character_)
  formatC(x, format = "g", digits = digits)
}

is_html_report <- function() isTRUE(knitr::is_html_output())

report_table <- function(data, rownames = FALSE, filter = "top",
                         extensions = NULL, options = list(),
                         caption = NULL, signif_cols = character(),
                         digits = 3) {
  data <- as.data.frame(data)
  if (is_html_report()) {
    dt_args <- list(
      data = data,
      rownames = rownames,
      filter = filter,
      caption = caption,
      options = options
    )
    if (!is.null(extensions)) dt_args$extensions <- extensions
    dt <- do.call(DT::datatable, dt_args)
    if (length(signif_cols) > 0) {
      dt <- DT::formatSignif(dt, columns = intersect(signif_cols, names(data)),
                             digits = digits)
    }
    return(dt)
  }

  static <- data
  for (col in intersect(signif_cols, names(static))) {
    if (is.numeric(static[[col]])) static[[col]] <- signif(static[[col]], digits)
  }
  max_rows <- options$pageLength %||% min(25, nrow(static))
  if (is.finite(max_rows) && nrow(static) > max_rows) {
    static <- utils::head(static, max_rows)
    caption <- paste(caption %||% "", sprintf("(showing first %d rows)", max_rows))
  }
  knitr::kable(static, caption = caption, booktabs = TRUE, longtable = TRUE)
}
```

::: {.panel-tabset .hidden}
:::

# Cover

```{r cover, results='asis'}
project_name <- project$name   %||% "Bulk RNA-seq Project"
analyst      <- project$analyst %||% "n/a"
report_date  <- format(Sys.Date(), "%Y-%m-%d")

cell_line <- exp_meta$cell_line %||% ""
compound  <- exp_meta$compound  %||% ""
dose      <- exp_meta$dose      %||% ""
duration  <- exp_meta$duration  %||% ""
notes     <- exp_meta$notes     %||% ""

empty_dash <- function(x) if (is.null(x) || is.na(x) || !nzchar(x)) "—" else x

cat("**Project:** ", project_name, "  \n", sep = "")
cat("**Prepared by:** ", analyst,  "  \n", sep = "")
cat("**Report date:** ", report_date, "\n\n", sep = "")

cat("## Experiment\n\n")
cat("| Field | Value |\n|---|---|\n")
cat(sprintf("| Cell line | %s |\n", empty_dash(cell_line)))
cat(sprintf("| Compound  | %s |\n", empty_dash(compound)))
cat(sprintf("| Dose      | %s |\n", empty_dash(dose)))
cat(sprintf("| Duration  | %s |\n", empty_dash(duration)))
cat(sprintf("| Notes     | %s |\n", empty_dash(notes)))
cat("\n")

cat("## Design\n\n")
cat(sprintf("- Samples: **%d**\n", nrow(samples)))
if ("condition" %in% names(samples)) {
  cond_tbl <- sort(table(samples$condition))
  cat("- Condition counts:\n")
  for (nm in names(cond_tbl)) {
    cat(sprintf("  - %s: %d\n", nm, cond_tbl[[nm]]))
  }
}
cat(sprintf("- Contrasts defined: **%d**\n", nrow(contrasts)))
cat("\n")
```

### Samples

```{r cover-samples}
knitr::kable(samples, row.names = FALSE, booktabs = TRUE, longtable = TRUE)
```

### Contrasts

```{r cover-contrasts}
knitr::kable(contrasts, row.names = FALSE, booktabs = TRUE, longtable = TRUE)
```

# Analysis workflow

```{mermaid}
flowchart TD
    A[nf-core/rnaseq · salmon counts] --> B[QC summary · MultiQC]
    A --> C[Exploratory · VST / PCA / sample correlation]
    C --> D[Differential expression · DESeq2 + apeglm]
    D --> E[Functional enrichment]
    E --> E1[GSEA · MSigDB]
    E --> E2[ORA · GO / KEGG / Reactome / Hallmark]
    D --> F[TFEA · decoupler ULM + CollecTRI]
    D --> G[Pathway activity · PROGENy]
    D --> H[cMap · L2S2 / LINCS]
```

## Report structure and interpretation

::: lang-en
| Stage | Analytical objective | Method | Primary deliverable |
|------------------|------------------|------------------|------------------|
| **Quality control** | Assess library preparation and alignment integrity. | MultiQC aggregation of FastQC, STAR/Salmon, and RSeQC metrics. | Per-sample table of alignment, contamination, complexity, and handling-failure indicators (5'/3' bias, intronic/intergenic, insert size, GC). |
| **Exploratory analysis** | Evaluate sample grouping structure and detect outliers prior to inference. | VST-transformed counts; PCA on top 500 most-variable genes; sample-to-sample correlation. | PCA projection, correlation heatmap, dendrogram. |
| **Differential expression** | Quantify condition-associated changes in gene expression. | DESeq2 Wald test with apeglm log-fold-change shrinkage. | Results table with shrunken LFC, Wald statistic, and adjusted *p*-values; volcano, MA, and top-DEG heatmap. |
| **Gene set enrichment (GSEA)** | Identify gene sets coordinately perturbed across the full ranked list. | Pre-ranked GSEA on Wald statistic across selected MSigDB collections. | Normalized enrichment score and adjusted *p*-value per gene set. |
| **Over-representation (ORA)** | Identify pathways enriched among the DEG shortlist. | Hypergeometric test against GO, KEGG, Reactome, and Hallmark databases. | Raw and adjusted *p*-values per pathway and database. |
| **TFEA** | Infer upstream regulators consistent with the observed expression changes. | decoupleR univariate linear model (ULM) applied to the CollecTRI regulon network. | Per-sample TF activity scores with significance. |
| **Pathway activity** | Estimate canonical signalling-pathway activity per contrast. | decoupleR multivariate linear model (MLM) on the DESeq2 Wald statistic against PROGENy with top responsive targets. | Per-contrast pathway activity scores with *p*-values. |
| **cMap** | Identify perturbagens producing transcriptomic signatures concordant or discordant with the contrast. | L2S2 / LINCS L1000 signature query using top up- and down-regulated genes. | Ranked perturbagen table with connectivity scores. |
| **Audit trail** | Ensure reproducibility and provenance of the analysis. | Snapshot of configuration, input checksums, and computational environment. | Resolved config, input MD5 sums, and full session information. |
:::

::: lang-ko
| 단계 | 분석 목적 | 방법 | 주 산출물 |
|------------------|------------------|------------------|------------------|
| **Quality control** | 라이브러리 준비와 정렬 품질 평가. | FastQC, STAR/Salmon, RSeQC metric을 MultiQC로 집계. | 샘플별 alignment, contamination, complexity, 처리 실패 지표 (5'/3' bias, intronic/intergenic, insert size, GC). |
| **탐색 분석** | 샘플 그룹 구조 확인 및 Outlier 확인. | VST 변환 카운트 · top 500 variable gene PCA · 샘플-샘플 Correlation. | PCA 투영, 상관 heatmap, dendrogram. |
| **차등발현** | 조건별 유전자 발현 변화 정량화. | DESeq2 Wald test + apeglm log-fold-change shrinkage. | shrunken LFC, Wald statistic, 조정 *p*-value 결과 표 · volcano · MA · top-DEG heatmap. |
| **Gene set enrichment (GSEA)** | 활성/비활성화 된 gene set 탐색. | 선택된 MSigDB collection에 대해 Wald statistic 기반 pre-ranked GSEA. | Gene set별 NES 및 조정 *p*-value. |
| **Over-representation (ORA)** | DEG 목록에서 과대표현된 pathway 확인. | GO, KEGG, Reactome, Hallmark DB에 대한 hypergeometric test. | Pathway·DB별 raw / adjusted *p*-value. |
| **TFEA** | 발현 패턴을 설명하는 상위 조절자 추론. | CollecTRI regulon 네트워크에 대한 decoupleR ULM. | 샘플별 TF 활성 score와 유의성. |
| **Pathway 활성도** | Contrast별 canonical signalling pathway 활성 추정. | DESeq2 Wald statistic을 PROGENy network에 대해 decoupleR MLM 적용 (top responsive targets). | Contrast별 pathway 활성 score와 *p*-value. |
| **cMap** | 해당 contrast와 일치하거나 반대되는 transcriptomic signature의 perturbagen 탐색. | Top up / down 유전자로 L2S2 / LINCS L1000 signature 질의. | Connectivity score 기반 perturbagen ranking 표. |
| **Audit trail** | 분석 재현성과 provenance 보장. | Config snapshot, 입력 checksum, 계산 환경 스냅샷. | 최종 config, 입력 MD5, 세션 정보. |
:::

{{< include sections/_qc.qmd >}}

{{< include sections/_exploratory.qmd >}}

{{< include sections/_de.qmd >}}

{{< include sections/_enrichment.qmd >}}

{{< include sections/_cmap.qmd >}}

# Audit trail

## Cutoffs in use

```{r audit-cutoffs, results='asis'}
de_cfg <- cfg$de %||% list()
prim   <- de_cfg$primary   %||% list(padj = NA, abs_lfc = NA)
sec    <- de_cfg$secondary %||% list(padj = NA, abs_lfc = NA)
ora_cfg <- cfg$enrichment$ora  %||% list()
gsea_cfg <- cfg$enrichment$gsea %||% list()
tf_cfg   <- cfg$tfea    %||% list()
dr_cfg   <- cfg$cmap    %||% list()

cat("| Setting | Value |\n|---|---|\n")
cat(sprintf("| Enrichment enabled | %s |\n", cfg$enrichment$enabled %||% "—"))
cat(sprintf("| TFEA enabled | %s |\n", cfg$tfea$enabled %||% "—"))
cat(sprintf("| PROGENy enabled | %s |\n", cfg$progeny$enabled %||% "—"))
cat(sprintf("| cMap enabled | %s |\n", cfg$cmap$enabled %||% "—"))
cat(sprintf("| Primary DEG | padj < %s & |LFC| >= %s |\n",
            prim$padj   %||% "—", prim$abs_lfc   %||% "—"))
cat(sprintf("| Secondary DEG | padj < %s & |LFC| >= %s |\n",
            sec$padj    %||% "—", sec$abs_lfc    %||% "—"))
cat(sprintf("| ORA min / max input genes | %s / %s |\n",
            ora_cfg$min_input_genes %||% "—", ora_cfg$max_input_genes %||% "—"))
cat(sprintf("| ORA databases | %s |\n",
            paste(vapply(ora_cfg$databases %||% list(),
                         function(d) d$label %||% d$id, character(1)),
                  collapse = ", ")))
cat(sprintf("| GSEA ranking | %s |\n",
            gsea_cfg$ranking %||% "—"))
cat(sprintf("| GSEA collections | %s |\n",
            paste(vapply(gsea_cfg$collections %||% list(),
                         function(c) c$label %||% c$id, character(1)),
                  collapse = ", ")))
cat(sprintf("| TFEA (min_size / top_n) | %s / %s |\n",
            tf_cfg$min_size %||% "—",
            tf_cfg$top_n %||% "—"))
cat(sprintf("| cMap (top_up / top_down) | %s / %s |\n",
            dr_cfg$top_up  %||% "—",
            dr_cfg$top_down %||% "—"))
cat("\n")
```

## Reproducibility summary

Pipeline source: <https://github.com/artblakey19/BulkRNAseq-Analyzer>

```{r audit-reproducibility}
file_meta <- function(label, path) {
  resolved <- if (is.null(path) || is.na(path) || !nzchar(path)) {
    NA_character_
  } else if (grepl("^/", path)) {
    path
  } else {
    file.path(project_root, path)
  }

  exists <- !is.na(resolved) && file.exists(resolved)
  info <- if (exists) file.info(resolved) else NULL
  md5 <- if (exists) unname(tools::md5sum(resolved)) else NA_character_

  data.frame(
    item = label,
    path = if (is.na(resolved)) NA_character_ else normalizePath(resolved, winslash = "/", mustWork = FALSE),
    exists = exists,
    size_bytes = if (exists) info$size else NA_real_,
    modified = if (exists) format(info$mtime, "%Y-%m-%d %H:%M:%S %Z") else NA_character_,
    md5 = md5,
    stringsAsFactors = FALSE
  )
}

# Pipeline version comes from the release git tag (the VERSION file was
# retired). Docker bakes it at build time into BULK_RNASEQ_VERSION (set from
# the tag in docker.yml); native/Colab clones fall back to `git describe`
# against the project tree.
pipeline_version <- local({
  env_ver <- Sys.getenv("BULK_RNASEQ_VERSION", unset = "")
  if (nzchar(env_ver) && !identical(env_ver, "unknown")) return(env_ver)
  git_ver <- tryCatch(
    suppressWarnings(system2(
      "git",
      c("-C", shQuote(project_root), "describe", "--tags", "--always", "--dirty"),
      stdout = TRUE, stderr = FALSE)),
    error = function(e) character(0))
  if (length(git_ver) && nzchar(git_ver[[1]])) git_ver[[1]] else "unknown"
})

run_meta <- data.frame(
  item = c("Report render time", "Pipeline version", "Project root", "Config used by report"),
  value = c(
    format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"),
    pipeline_version,
    project_root,
    config_path
  ),
  stringsAsFactors = FALSE
)

knitr::kable(run_meta, row.names = FALSE, booktabs = TRUE, longtable = TRUE)

# nf-core/rnaseq provenance files (optional). Resolve the latest params_*.json
# and the software-versions YAML so the upstream aligner/quantifier choice and
# tool versions are captured here rather than re-typed into a free-text note.
# Globs are sorted; the lexically-last params file is the most recent run
# (timestamped names), which is the one whose counts this report consumed.
nfcore_dir <- cfg$input$nfcore_pipeline_info_dir %||% ""
nfcore_params_path   <- NA_character_
nfcore_versions_path <- NA_character_
if (nzchar(nfcore_dir)) {
  nfcore_dir_abs <- if (grepl("^(/|~|[A-Za-z]:[/\\\\])", nfcore_dir)) nfcore_dir else rp(nfcore_dir)
  params_hits <- sort(Sys.glob(file.path(nfcore_dir_abs, "params_*.json")))
  if (length(params_hits)) nfcore_params_path <- params_hits[length(params_hits)]
  ver_hits <- Sys.glob(file.path(nfcore_dir_abs, "*software*versions*.yml"))
  if (length(ver_hits)) nfcore_versions_path <- ver_hits[1]
}

input_meta <- dplyr::bind_rows(
  file_meta("Config YAML", config_path),
  file_meta("Counts TSV", cfg$input$counts_tsv %||% NA_character_),
  file_meta("Samples TSV", cfg$input$samples_tsv %||% NA_character_),
  file_meta("Contrasts TSV", cfg$input$contrasts_tsv %||% NA_character_),
  file_meta("MultiQC data directory", cfg$input$multiqc_data_dir %||% NA_character_)
)

if (nzchar(nfcore_dir)) {
  input_meta <- dplyr::bind_rows(
    input_meta,
    file_meta("nf-core params JSON", nfcore_params_path),
    file_meta("nf-core software versions", nfcore_versions_path)
  )
}

knitr::kable(input_meta, row.names = FALSE, booktabs = TRUE, longtable = TRUE)
```

## nf-core/rnaseq provenance

```{r audit-nfcore-provenance, results='asis'}
if (is.na(nfcore_params_path) && is.na(nfcore_versions_path)) {
  cat(as.character(no_results(paste(
    "No nf-core/rnaseq pipeline_info directory configured",
    "(input.nfcore_pipeline_info_dir) — the upstream aligner/quantifier choice",
    "and tool versions are not captured in this report."
  ))))
} else {
  # Curated params: keys that change the quantity downstream analysis consumes.
  # aligner vs pseudo_aligner is the STAR/Salmon split the analyst cares about.
  if (!is.na(nfcore_params_path)) {
    nf_params <- jsonlite::fromJSON(nfcore_params_path, simplifyVector = TRUE)
    keys <- c(
      aligner              = "Aligner",
      pseudo_aligner       = "Pseudo-aligner",
      genome               = "iGenomes key",
      fasta                = "Genome FASTA",
      gtf                  = "Annotation GTF",
      gtf_extra_attributes = "GTF extra attributes",
      gencode              = "GENCODE mode",
      trimmer              = "Trimmer",
      with_umi             = "UMI handling"
    )
    present <- keys[names(keys) %in% names(nf_params)]
    fmt <- function(v) if (is.null(v) || length(v) == 0) "—" else paste(format(v), collapse = ", ")
    param_tbl <- data.frame(
      parameter = unname(present),
      value     = vapply(names(present), function(k) fmt(nf_params[[k]]), character(1)),
      stringsAsFactors = FALSE
    )
    cat(sprintf("**Upstream pipeline parameters** (`%s`)\n\n", basename(nfcore_params_path)))
    cat(knitr::kable(param_tbl, row.names = FALSE, booktabs = TRUE, longtable = TRUE), sep = "\n")
    cat("\n\n")
  }

  # Full version table flattened from the per-process YAML. The Workflow block
  # carries the pipeline revision + Nextflow version.
  if (!is.na(nfcore_versions_path)) {
    vers <- yaml::read_yaml(nfcore_versions_path)
    rows <- lapply(names(vers), function(proc) {
      tools <- vers[[proc]]
      data.frame(
        process = proc,
        tool    = names(tools),
        version = vapply(tools, function(x) paste(format(x), collapse = ", "), character(1)),
        stringsAsFactors = FALSE
      )
    })
    ver_tbl <- do.call(rbind, rows)
    rownames(ver_tbl) <- NULL
    cat(sprintf("**Software versions** (`%s`)\n\n", basename(nfcore_versions_path)))
    cat(knitr::kable(ver_tbl, row.names = FALSE, booktabs = TRUE, longtable = TRUE), sep = "\n")
  }
}
```

## Tools and references

**Workflow infrastructure**

-   Snakemake — Mölder F. et al. (2021). *F1000Research* 10:33. [doi:10.12688/f1000research.29032.2](https://doi.org/10.12688/f1000research.29032.2)
-   nf-core/rnaseq — Ewels P.A. et al. (2020). *Nat Biotechnol.* 38, 276–278. [doi:10.1038/s41587-020-0439-x](https://doi.org/10.1038/s41587-020-0439-x)
-   Quarto — <https://quarto.org>

**Quantification, counts, and QC**

-   Salmon — Patro R. et al. (2017). *Nat Methods* 14, 417–419. [doi:10.1038/nmeth.4197](https://doi.org/10.1038/nmeth.4197)
-   tximport — Soneson C., Love M.I., Robinson M.D. (2015). *F1000Research* 4:1521. [doi:10.12688/f1000research.7563.2](https://doi.org/10.12688/f1000research.7563.2)
-   MultiQC — Ewels P. et al. (2016). *Bioinformatics* 32(19), 3047–3048. [doi:10.1093/bioinformatics/btw354](https://doi.org/10.1093/bioinformatics/btw354)

**Differential expression**

-   DESeq2 — Love M.I., Huber W., Anders S. (2014). *Genome Biol.* 15, 550. [doi:10.1186/s13059-014-0550-8](https://doi.org/10.1186/s13059-014-0550-8)
-   apeglm — Zhu A., Ibrahim J.G., Love M.I. (2019). *Bioinformatics* 35(12), 2084–2092. [doi:10.1093/bioinformatics/bty895](https://doi.org/10.1093/bioinformatics/bty895)

**Gene-set analysis**

-   GSEA (method) — Subramanian A. et al. (2005). *PNAS* 102(43), 15545–15550. [doi:10.1073/pnas.0506580102](https://doi.org/10.1073/pnas.0506580102)
-   fgsea — Korotkevich G. et al. (2021). *bioRxiv.* [doi:10.1101/060012](https://doi.org/10.1101/060012)
-   clusterProfiler 4.0 — Wu T. et al. (2021). *The Innovation* 2(3), 100141. [doi:10.1016/j.xinn.2021.100141](https://doi.org/10.1016/j.xinn.2021.100141)
-   MSigDB / Hallmark — Liberzon A. et al. (2015). *Cell Systems* 1(6), 417–425. [doi:10.1016/j.cels.2015.12.004](https://doi.org/10.1016/j.cels.2015.12.004)

**Regulatory and pathway activity inference**

-   decoupleR / decoupler-py — Badia-i-Mompel P. et al. (2022). *Bioinformatics Advances* 2(1), vbac016. [doi:10.1093/bioadv/vbac016](https://doi.org/10.1093/bioadv/vbac016)
-   CollecTRI — Müller-Dott S. et al. (2023). *Nucleic Acids Research* gkad841. [doi:10.1093/nar/gkad841](https://doi.org/10.1093/nar/gkad841)
-   PROGENy — Schubert M. et al. (2018). *Nat Commun* 9, 20. [doi:10.1038/s41467-017-02391-6](https://doi.org/10.1038/s41467-017-02391-6)

**cMap / connectivity**

-   L1000 Connectivity Map — Subramanian A. et al. (2017). *Cell* 171(6), 1437–1452. [doi:10.1016/j.cell.2017.10.049](https://doi.org/10.1016/j.cell.2017.10.049)
-   L2S2 — L1000 signature search (Ma'ayan Lab). <https://l2s2.maayanlab.cloud>

## Session info

```{r audit-sessioninfo}
# Keep only reproducibility-relevant fields: R version, platform, OS,
# attached packages + versions, and loaded-via-namespace packages.
# Drop: BLAS/LAPACK install paths, locale, time zone, RNG kind.
si <- utils::capture.output(sessionInfo())

drop_pat <- paste0(
  "^(Matrix products:",
  "|BLAS(/LAPACK)?:|LAPACK:",
  "|locale:",
  "|time zone:",
  "|tzcode source:",
  "|Random number generation:",
  "|\\s+\\[\\s*\\d+\\]\\s+LC_",
  "|\\s+(RNG|Normal|Sample):)"
)
si <- si[!grepl(drop_pat, si)]

# Collapse runs of consecutive blank lines left behind by removals.
prev_blank <- FALSE
keep <- logical(length(si))
for (i in seq_along(si)) {
  is_blank <- !nzchar(trimws(si[i]))
  keep[i] <- !(is_blank && prev_blank)
  prev_blank <- is_blank
}
cat(si[keep], sep = "\n")
```