# Functional enrichment

::: lang-en
Per-contrast GSEA, ORA, TFEA (CollecTRI / ULM), and PROGENy pathway activity.
:::

::: lang-ko
Contrast별 GSEA, ORA, TFEA (CollecTRI / ULM), PROGENy pathway 활성도.
:::

````{r enr-child-template, include=FALSE}
enr_child_template <- r"---(
## `r cid`

### GSEA

```{r enr-gsea, results='asis', fig.width=9, fig.height=8}
gsea_path <- file.path(project_root, "results", "enrichment", cid, "gsea_combined.csv")
gsea_all <- safe_read(gsea_path)

.coll_entries <- cfg$enrichment$gsea$collections %||% list()
.coll_labels <- setNames(
  vapply(.coll_entries, function(x) x$label %||% x$id, character(1)),
  vapply(.coll_entries, function(x) x$id, character(1))
)
.collection_label <- function(id) {
  out <- unname(.coll_labels[id])
  ifelse(is.na(out), id, out)
}

.trunc_label <- function(x, n = 50) {
  x <- as.character(x)
  ifelse(nchar(x) > n, paste0(substr(x, 1, n - 1), "..."), x)
}

.top_by_direction <- function(df, direction_col = "direction", p_col = "padj", n = 10) {
  if (!all(c(direction_col, p_col) %in% names(df))) return(df[0, , drop = FALSE])
  df |>
    dplyr::filter(.data[[direction_col]] %in% c("up", "down")) |>
    dplyr::group_by(.data[[direction_col]]) |>
    dplyr::arrange(.data[[p_col]], .by_group = TRUE) |>
    dplyr::slice_head(n = n) |>
    dplyr::ungroup()
}

if (!isTRUE(enrichment_enabled)) {
  cat(as.character(no_results("GSEA / ORA stage disabled in config.yaml.")))
} else if (is.null(gsea_all) || nrow(gsea_all) == 0) {
  cat(as.character(no_results("GSEA returned no enriched sets for this contrast.")))
} else {
  colls <- unique(as.character(gsea_all$collection))
  colls <- colls[!is.na(colls) & nzchar(colls)]
  if (is_html_report() && length(colls) > 0) {
    cat("\n\n::: {.panel-tabset}\n\n")
  }
  for (coll in colls) {
    df <- gsea_all |> dplyr::filter(collection == coll)
    if (nrow(df) == 0) next
    grp <- .collection_label(coll)
    cat(sprintf("\n\n#### %s\n\n", grp))
    ranked <- df |>
      dplyr::mutate(direction = dplyr::case_when(
        NES > 0 ~ "up",
        NES < 0 ~ "down",
        TRUE ~ NA_character_
      ))
    top <- .top_by_direction(ranked, "direction", "padj", n = 10) |>
      dplyr::mutate(label = .trunc_label(pathway))
    top$label <- factor(top$label, levels = rev(unique(top$label)))
    p <- ggplot2::ggplot(top, ggplot2::aes(x = NES, y = label, fill = direction)) +
      ggplot2::geom_col() +
      ggplot2::scale_fill_manual(values = c(up = "#c0392b", down = "#2c7bb6")) +
      ggplot2::geom_vline(xintercept = 0, colour = "grey40") +
      ggplot2::labs(x = "NES", y = NULL,
                    title = sprintf("Top 10 up / Top 10 down %s gene sets (by padj)", grp)) +
      ggplot2::theme_minimal(base_size = 11) +
      ggplot2::theme(
        axis.text.y = ggplot2::element_text(size = 9),
        plot.margin = ggplot2::margin(6, 10, 6, 6)
      )
    print(p)
    cat("\n\n")

    show_df <- ranked |>
      .top_by_direction("direction", "padj", n = 10) |>
      dplyr::select(dplyr::any_of(c("collection", "pathway", "NES",
                                    "pval", "padj", "size", "leadingEdge")))
    dt <- report_table(
      show_df,
      rownames = FALSE,
      filter   = "top",
      options  = list(
        pageLength = 20,
        lengthMenu = c(10, 20, 50, 100),
        scrollX = TRUE,
        autoWidth = FALSE,
        dom = "lfrtip",
        columnDefs = list(
          list(targets = which(names(show_df) == "pathway") - 1,
               className = "pathway-name-column")
        )
      ),
      signif_cols = intersect(c("NES", "pval", "padj"), names(show_df)),
      digits  = 3
    )
    cat(knitr::knit_print(dt))
    cat("\n\n")
  }
  if (is_html_report() && length(colls) > 0) {
    cat("\n\n:::\n\n")
  }
}
```

### ORA (supplementary)

```{r enr-ora, results='asis', fig.width=9, fig.height=11}
ora_min_genes <- cfg$enrichment$ora$min_input_genes %||% 100
ora_input_path <- file.path(project_root, "results", "enrichment", cid, "ora_input_genes.csv")
ora_input_genes <- safe_read(ora_input_path)
ora_input_cutoff <- if (!is.null(ora_input_genes) && nrow(ora_input_genes) > 0)
                      unique(ora_input_genes$cutoff)[1] else NA_character_

if (is_html_report()) {
  cat("\n\n<details>\n")
  cat("<summary><strong>ORA input gene list</strong></summary>\n\n")
} else {
  cat("\n\n#### ORA input gene list\n\n")
}
if (!isTRUE(enrichment_enabled)) {
  cat(as.character(no_results("GSEA / ORA stage disabled in config.yaml.")))
  cat("\n\n")
} else if (is.null(ora_input_genes) || nrow(ora_input_genes) == 0) {
  cat(as.character(no_results(sprintf(
    "No ORA input gene list selected: primary and secondary cutoffs both yielded fewer than %d genes.",
    ora_min_genes
  ))))
  cat("\n\n")
} else {
  gene_counts <- table(factor(ora_input_genes$direction, levels = c("up", "down")))
  caption <- sprintf(
    "ORA input genes: %s cutoff, %d total (%d up, %d down)",
    ora_input_cutoff,
    nrow(ora_input_genes),
    as.integer(gene_counts[["up"]] %||% 0),
    as.integer(gene_counts[["down"]] %||% 0)
  )
  dt <- report_table(
    ora_input_genes,
    rownames = FALSE,
    filter = "top",
    caption = caption,
    options = list(
      pageLength = 25,
      lengthMenu = c(25, 50, 100, 250, 500),
      scrollX = TRUE,
      autoWidth = FALSE,
      dom = "lfrtip"
    ),
    signif_cols = intersect(c("log2FoldChange", "padj"), names(ora_input_genes)),
    digits = 3
  )
  cat(knitr::knit_print(dt))
  cat("\n\n")
}
if (is_html_report()) cat("\n</details>\n\n")

cat("\n\n#### ORA enrichment results\n\n")
ora_path <- file.path(project_root, "results", "enrichment", cid, "ora_combined.csv")
ora <- safe_read(ora_path)

.db_entries <- cfg$enrichment$ora$databases %||% list()
.db_labels <- setNames(
  vapply(.db_entries, function(x) x$label %||% x$id, character(1)),
  vapply(.db_entries, function(x) x$id, character(1))
)
.database_label <- function(id) {
  out <- unname(.db_labels[id])
  ifelse(is.na(out), id, out)
}

if (!isTRUE(enrichment_enabled)) {
  cat(as.character(no_results("GSEA / ORA stage disabled in config.yaml.")))
} else if (is.null(ora) || nrow(ora) == 0) {
  cat(as.character(no_results("ORA returned no results (insufficient DEG or empty output).")))
} else {
  ora_dbs <- unique(as.character(ora$database))
  ora_dbs <- ora_dbs[!is.na(ora_dbs) & nzchar(ora_dbs)]

  if (is_html_report() && length(ora_dbs) > 0) {
    cat("\n\n::: {.panel-tabset}\n\n")
  }

  for (db in ora_dbs) {
    ora_db <- ora |> dplyr::filter(database == db)
    if (nrow(ora_db) == 0) next
    db_lbl <- .database_label(db)
    cat(sprintf("\n\n##### %s\n\n", db_lbl))

    show_df <- ora_db |>
      dplyr::filter(direction %in% c("up", "down")) |>
      dplyr::group_by(direction) |>
      dplyr::arrange(p.adjust, .by_group = TRUE) |>
      dplyr::slice_head(n = 10) |>
      dplyr::ungroup() |>
      dplyr::select(dplyr::any_of(c("direction", "cutoff", "ID",
                                    "Description", "GeneRatio", "BgRatio",
                                    "pvalue", "p.adjust", "qvalue", "Count")))

    bars <- show_df |>
      dplyr::mutate(
        neglog10p = -log10(pmax(p.adjust, 1e-300)),
        Description = ifelse(is.na(Description) | !nzchar(as.character(Description)),
                             as.character(ID), as.character(Description)),
        Description = ifelse(nchar(Description) > 50,
                             paste0(substr(Description, 1, 49), "..."),
                             Description)
      )
    if (nrow(bars) > 0) {
      p <- ggplot2::ggplot(
          bars,
          ggplot2::aes(x = neglog10p,
                       y = reorder(Description, neglog10p),
                       fill = direction)) +
        ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.8)) +
        ggplot2::scale_fill_manual(values = c(up = "#c0392b", down = "#2c7bb6")) +
        ggplot2::labs(x = "-log10(p.adjust)", y = NULL,
                      title = sprintf("Top 10 up / Top 10 down %s terms (by p.adjust)", db_lbl)) +
        ggplot2::theme_minimal(base_size = 11) +
        ggplot2::theme(
          axis.text.y = ggplot2::element_text(size = 9),
          plot.margin = ggplot2::margin(6, 15, 6, 15)
        )
      print(p)
      cat("\n\n")
    }

    dt <- report_table(
      show_df,
      rownames = FALSE,
      filter   = "top",
      options  = list(
        pageLength = 20,
        lengthMenu = c(10, 20, 50, 100),
        scrollX = TRUE,
        autoWidth = FALSE,
        dom = "lfrtip",
        columnDefs = list(
          list(targets = which(names(show_df) == "Description") - 1,
               className = "pathway-name-column")
        )
      ),
      signif_cols = intersect(c("pvalue", "p.adjust", "qvalue"), names(show_df)),
      digits  = 3
    )
    cat(knitr::knit_print(dt))
    cat("\n\n")
  }

  if (is_html_report() && length(ora_dbs) > 0) {
    cat("\n\n:::\n\n")
  }
}
```

### TFEA (CollecTRI, ULM)

```{r enr-tf-activity, results='asis', fig.width=8, fig.height=9}
tf_path <- file.path(project_root, "results", "tfea", cid, "tf_scores.tsv")
tf <- safe_read(tf_path, sep = "\t")
if (!isTRUE(tfea_enabled)) {
  cat(as.character(no_results("TFEA stage disabled in config.yaml.")))
} else if (is.null(tf) || nrow(tf) == 0) {
  cat(as.character(no_results("TFEA output is empty (network fetch may have failed).")))
} else {
  tf_top <- tf |>
    dplyr::mutate(abs_s = abs(score)) |>
    dplyr::arrange(dplyr::desc(abs_s)) |>
    head(30) |>
    dplyr::mutate(source = factor(source, levels = rev(source)))
  p <- ggplot2::ggplot(tf_top,
          ggplot2::aes(x = score, y = source,
                       fill = ifelse(score > 0, "up", "down"))) +
    ggplot2::geom_col() +
    ggplot2::scale_fill_manual(values = c(up = "#c0392b", down = "#2c7bb6"),
                               name = "direction") +
    ggplot2::geom_vline(xintercept = 0, colour = "grey40") +
    ggplot2::labs(x = "ULM score", y = NULL,
                  title = "Top 30 TFs by |score|") +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      axis.text.y = ggplot2::element_text(size = 9),
      plot.margin = ggplot2::margin(6, 10, 6, 6)
    )
  print(p)
  cat("\n\n")

  show_df <- tf |>
    dplyr::arrange(dplyr::desc(abs(score))) |>
    head(200)
  dt <- report_table(
    show_df, rownames = FALSE, filter = "top",
    options  = list(pageLength = 10, lengthMenu = c(10, 25, 50, 100), scrollX = TRUE, dom = "lfrtip"),
    signif_cols = intersect(c("score", "p_value"), names(show_df)),
    digits  = 3
  )
  cat(knitr::knit_print(dt))
  cat("\n\n")
}
```

### Pathway activity (PROGENy)

::: {.lang-en}
Contrast-level pathway activity inferred by decoupler's MLM on the DESeq2 Wald statistic against the PROGENy network. Positive score → pathway up in numerator; negative → up in denominator.
:::

::: {.lang-ko}
DESeq2 Wald statistic을 PROGENy network에 대해 decoupler의 MLM으로 contrast 단위 pathway 활성도 추정. score > 0이면 numerator에서 활성, < 0이면 denominator에서 활성.
:::

```{r enr-progeny-bars, fig.width=7.5, fig.height=6}
pw_path <- file.path(project_root, "results", "progeny", cid,
                     "progeny_scores.tsv")
pw <- safe_read(pw_path, sep = "\t")

if (!isTRUE(progeny_enabled)) {
  no_results("PROGENy stage disabled in config.yaml.")
} else if (is.null(pw) || nrow(pw) == 0 ||
    !all(c("pathway", "score") %in% names(pw)) ||
    all(is.na(pw$score))) {
  no_results("PROGENy scores unavailable or all-NA.")
} else {
  df <- pw |>
    dplyr::filter(!is.na(score)) |>
    dplyr::mutate(direction = ifelse(score > 0, "up", "down"),
                  pathway   = factor(pathway, levels = pathway[order(score)]))

  p <- ggplot2::ggplot(df, ggplot2::aes(x = score, y = pathway, fill = direction)) +
    ggplot2::geom_col() +
    ggplot2::scale_fill_manual(values = c(up = "#c0392b", down = "#2c7bb6"),
                               guide = "none") +
    ggplot2::geom_vline(xintercept = 0, colour = "grey40") +
    ggplot2::labs(x = "MLM activity score", y = NULL,
                  title = "PROGENy pathway activity",
                  subtitle = sprintf("%s — input: DESeq2 Wald stat", cid)) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(axis.text.y = ggplot2::element_text(size = 10))
  p
}
```

```{r enr-progeny-table}
if (!isTRUE(progeny_enabled)) {
  no_results("PROGENy stage disabled in config.yaml.")
} else if (is.null(pw) || nrow(pw) == 0 ||
    !all(c("pathway", "score") %in% names(pw))) {
  no_results("PROGENy table unavailable.")
} else {
  show_df <- pw |>
    dplyr::arrange(dplyr::desc(abs(score)))

  dt <- report_table(
    show_df, rownames = FALSE,
    caption  = sprintf("PROGENy activity (decoupler MLM on Wald stat) — %s", cid),
    options  = list(pageLength = 14, lengthMenu = c(14, 25, 50, 100),
                    dom = "lfrtip", scrollX = TRUE),
    signif_cols = intersect(c("score", "p_value", "padj"), names(show_df)),
    digits  = 3
  )
  dt
}
```


)---"
````

```{r enr-render, results='asis'}
if (nrow(contrasts) == 0) {
  print(no_results("No contrasts defined in contrasts.tsv."))
}

enrichment_enabled <- isTRUE(cfg$enrichment$enabled %||% TRUE)
tfea_enabled <- isTRUE(cfg$tfea$enabled %||% TRUE)
progeny_enabled <- isTRUE(cfg$progeny$enabled %||% TRUE)

for (cid in contrasts$contrast_id) {
  env <- new.env(parent = globalenv())
  env$cid                <- cid
  env$project_root       <- project_root
  env$safe_read          <- safe_read
  env$no_results         <- no_results
  env$enrichment_enabled <- enrichment_enabled
  env$tfea_enabled       <- tfea_enabled
  env$progeny_enabled    <- progeny_enabled
  env$`%||%`             <- `%||%`

  out <- knitr::knit_child(
    text  = enr_child_template,
    envir = env,
    quiet = TRUE
  )
  cat(out, sep = "\n")
}
```