# QC

::: lang-en
Per-sample QC metrics from MultiQC. Sample quality (degradation, contamination, library-prep deviations).
:::

::: lang-ko
MultiQC 출력에서 집계한 샘플별 QC 지표. 샘플 품질을(분해, 오염, 라이브러리 이상) 평가.
:::

```{r qc-load}
qc_path <- rp("results", "qc", "qc_summary.tsv")
qc <- safe_read(qc_path, sep = "\t")
```

```{r qc-meta}
# Single source of truth for the QC key-metric panel: order here sets facet
# order, and the description is rendered only for metrics actually present
# (non-NA). Salmon and STAR+RSeQC runs therefore show whichever subset they
# produce — the description can never name a metric the plot doesn't draw.
# Detail-only metrics (insert size, GC, read distribution, salmon ratios, …)
# stay in the per-sample table above, not the plot.
qc_meta <- list(
  mapping_rate = list(
    en = "Percentage of reads assigned to the reference. Low values may indicate contamination, poor read quality, or reference mismatch.",
    ko = "read가 reference에 할당된 비율임. 낮다면 contamination, read quality 문제, reference mismatch 등이 의심됨."
  ),

  duplication = list(
    en = "Percentage of duplicated reads. High duplication often suggests low library complexity, over-amplification, or low input material.",
    ko = "중복 read의 비율임. 높으면 library complexity가 낮거나, PCR 과증폭 또는 input 부족 가능성이 있음."
  ),

  rrna_pct = list(
    en = "Percentage of reads coming from rRNA. High values suggest inefficient rRNA removal or weak mRNA enrichment.",
    ko = "rRNA에서 유래한 read의 비율. 높으면 rRNA 제거가 잘 되지 않았거나 mRNA enrichment가 부족했을 수 있음."
  ),

  gene_body_5_3_bias = list(
    en = "5' to 3' coverage balance across gene bodies. Low values suggest 3' bias, commonly seen with degraded RNA.",
    ko = "gene body에서 5'와 3' coverage가 얼마나 균형적인지에 대한 지표. 낮으면 RNA degradation에서 흔한 3' bias가 의심됨."
  )
)

# Key metrics present as a column and not entirely NA, in qc_meta order.
qc_present <- function() {
  if (is.null(qc) || nrow(qc) == 0) return(character(0))
  m <- intersect(names(qc_meta), names(qc))
  m[vapply(m, function(c) any(!is.na(qc[[c]])), logical(1))]
}

# Render the bilingual bullet glossary for a set of metrics (asis markdown).
emit_glossary <- function(metrics) {
  if (length(metrics) == 0) return(invisible())
  for (lang in c("en", "ko")) {
    cat(sprintf("\n::: lang-%s\n\n", lang))
    for (m in metrics) cat(sprintf("- `%s` — %s\n", m, qc_meta[[m]][[lang]]))
    cat("\n:::\n")
  }
}
```

```{r qc-table}
if (is.null(qc) || nrow(qc) == 0) {
  no_results("QC summary table not available.")
} else {
  qc_show <- qc
  numeric_cols <- c("mapping_rate", "uniquely_mapped_pct",
                    "compatible_fragment_ratio", "strand_mapping_bias",
                    "rrna_pct", "duplication", "dupradar_slope",
                    "exonic_pct", "intronic_pct", "intergenic_pct",
                    "gene_body_5_3_bias", "insert_size_median",
                    "gc_content_pct")
  for (col in intersect(numeric_cols, names(qc_show))) {
    if (is.numeric(qc_show[[col]])) {
      qc_show[[col]] <- signif(qc_show[[col]], 3)
    }
  }
  knitr::kable(qc_show,
               row.names = FALSE,
               booktabs  = TRUE,
               longtable = TRUE,
               caption   = "Per-sample QC metrics")
}
```

## Library size

```{r qc-libsize, fig.width=7, fig.height=3.5}
if (is.null(qc) || nrow(qc) == 0 || !"assigned_reads" %in% names(qc)) {
  no_results("Per-sample assigned-read totals not available.")
} else {
  df <- qc |>
    dplyr::mutate(
      sample    = factor(sample, levels = sample),
      condition = as.character(condition)
    )
  p <- ggplot2::ggplot(df, ggplot2::aes(
        x = sample, y = assigned_reads, fill = condition
      )) +
    ggplot2::geom_col() +
    ggplot2::scale_y_continuous(labels = scales::comma) +
    ggplot2::labs(x = NULL, y = "Assigned reads (sum of counts)") +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
  print(p)
}
```

## Key QC metrics

::: lang-en
Core QC metrics for checking sample quality. Cohort-level outliers may indicate sample-specific problems.
:::

::: lang-ko
샘플 품질을 확인하기 위한 주요 QC 지표. Cohort 내 outlier가 있을 시 해당 샘플의 품질이나 처리 과정에 문제가 있었는지 점검요망.
:::

```{r qc-key-desc, results='asis'}
emit_glossary(qc_present())
```

```{r qc-key, fig.width=8, fig.height=5}
key_cols <- qc_present()
if (length(key_cols) == 0) {
  no_results("Key QC metrics not available.")
} else {
  df <- qc |>
    dplyr::select(sample, condition, dplyr::all_of(key_cols)) |>
    tidyr::pivot_longer(dplyr::all_of(key_cols),
                        names_to = "metric", values_to = "value") |>
    dplyr::mutate(sample = factor(sample, levels = unique(sample)),
                  metric = factor(metric, levels = key_cols))
  p <- ggplot2::ggplot(df, ggplot2::aes(x = sample, y = value, fill = condition)) +
    ggplot2::geom_col() +
    ggplot2::facet_wrap(~ metric, scales = "free_y") +
    ggplot2::labs(x = NULL, y = NULL) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
  print(p)
}
```

```{r qc-strand-check, results='asis'}
if (!is.null(qc) && "strandedness" %in% names(qc) && any(!is.na(qc$strandedness))) {
  calls <- unique(stats::na.omit(qc$strandedness))
  if (length(calls) == 1) {
    cat(sprintf(
      "_Inferred strandedness (RSeQC infer_experiment or Salmon library type): **%s** for all samples._\n",
      calls
    ))
  } else {
    tab <- table(qc$strandedness)
    mixed <- paste(sprintf("%s (n=%d)", names(tab), as.integer(tab)),
                   collapse = ", ")
    cat(sprintf(
      "_Inferred strandedness varies across samples: %s. Confirm library-preparation records before counting._\n",
      mixed
    ))
  }
}
```